---
title: "Lesson 10. Other bits and pieces"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "C:/Matias/Cursos/2019_Mexico/htmls") })

output: 
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: yes
    theme: united
---

<style type="text/css">
h1.title {
  font-size: 30px;
    color: Maroon;
  text-align: center;
          }
h3.subtitle {
  font-size: 22px;
    color: Maroon;
  text-align: center;
            }
h4.author { 
    font-size: 24px;
      color: FireBrick;
  text-align: center;
          }
h4.date { 
  font-size: 18px;
  text-align: center;
        }
h1{
  font-size: 22px;
  color: DarkBlue;
  }
h2{
  font-size: 20px;
  color: Blue;
  }
h3{
  font-size: 18px;
  color: SteelBlue;
  }
body{
    font-family: Helvetica;
    font-size: 14pt;
    }
code.r{
  font-size: 16pt;
      }
pre {
  font-size: 16pt;
}
</style>


```{r globaloptions, include=FALSE}
knitr::opts_chunk$set(fig.width = 6,fig.height = 6,
                        echo = TRUE, warning=FALSE,message=FALSE)
```
# Lesson goals
We're going to a range of other R related concepts and applications:
* Directories and file paths
* Vectorization (add purr package, function map, apply equivalent of tidyverse)
* Version control and Rmarkdown
* Building websites  (shiny, blogdown package) 

<br>
Credits to John Hoenig, Ainslie Denham and Alex Hesp
<br><br>


<br>

# Directories and file paths


<br><br>

# Vectorization



<br><br>

package here for managing paths and wd
https://malco.io/2018/11/05/why-should-i-use-the-here-package/


# Version control and Rmarkdown

add reference manager, equations, etc

What problems can you see with the following approach to version control?

- papersims-v1.R
- papersims-v2.R
- papersims-thea-comments.R
- papersims-hellfire.R
- papersims-v5.R
- papersims-final.R


# Version control

1. How many versions until this becomes intractable?
2. *Date Modified* sorting does not always help
3. Tracking changes is very difficult; reversion is even harder
4. Exponential file number growth with multiple collaborators!
5. The dreaded computer crash

Dropbox can help with some of these issues, but generally not (3) or (4)!

# Git

- git is an open source version control system (VCS):
    - Track changes to code and documents. *What* changes by *who* and *when*?
    - Share code and collaborate
- GitHub is a website that uses git's VCS:
    - Collaborate with others effectively
    - Distribute code
    - Solicit improvements (*pull* requests)
    - Track issues & feature requests
    - (Build your coding portfolio!)

# Git with R

git & GitHub are popular with R developers

- Integration with RStudio
- Easy distribution of packages
    - Circumvents CRAN moderators; for better or worse
- Always get the latest features (`devtools`; `install_github`)

# Git with life

- Collaborations: syncing across multiple computers
- Portability: including for you
- TODOs as Issues: including your own
- Teams: `fork`, `pull` request
    - Like a package but hate the default figures? `fork` it, edit it, use your own version!
- Visibility: social/professional network

# Getting started with git

- Download/update: [git-scm.com/downloads](git-scm.com/downloads)
- Intro: [guides.github.com/activities/hello-world](guides.github.com/activities/hello-world)
- Do this with Homework 1 and a blank pdf, not hello-world
- Questions? Error messages? The internet is a great resource!
- *git is a great habit to get into early! Start now!*

Homework 1 Question 0 will get you started using git, GitHub and GitHub Classroom.

# Getting started with git

The workflow that I want you to follow for downloading the latest lecture notes is `git pull` (see Question 0 of Homework 1)

The standard workflow for uploading a new file or updating an old one (e.g., for your homework submission):

```{r, eval = FALSE}
git pull
git add hw1-response.pdf
git commit -a -m 'question 2 part b response'
git push
```