---
title: "Lesson 7. Multivariate stats"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "C:/Matias/Cursos/2019_Mexico/htmls") })

output: 
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: yes
    theme: united
---

<style type="text/css">
h1.title {
  font-size: 30px;
    color: Maroon;
  text-align: center;
          }
h3.subtitle {
  font-size: 22px;
    color: Maroon;
  text-align: center;
            }
h4.author { 
    font-size: 24px;
      color: FireBrick;
  text-align: center;
          }
h4.date { 
  font-size: 18px;
  text-align: center;
        }
h1{
  font-size: 22px;
  color: DarkBlue;
  }
h2{
  font-size: 20px;
  color: Blue;
  }
h3{
  font-size: 18px;
  color: SteelBlue;
  }
body{
    font-family: Helvetica;
    font-size: 14pt;
    }
code.r{
  font-size: 16pt;
      }
pre {
  font-size: 16pt;
}
</style>


```{r globaloptions, include=FALSE}
knitr::opts_chunk$set(fig.width = 6,fig.height = 6,
                        echo = TRUE, warning=FALSE,message=FALSE)
```

# Lesson goals
We're going to apply a range of basic multivariate statistics.

This will not be a comprehensive analysis

Rather, it is a guide for those interested in learning how to implement this type of analyses in R

Much more (and better) can be done than what is shown here...



<br>
source
https://peat-clark.github.io/BIO381/veganTutorial.html

https://chrischizinski.github.io/rstats/adonis/

<br>
```{r}
library(vegan)

#1. load data
data(dune) # Vegetation and Environment in Dutch Dune Meadows
str(dune) #a data frame of observations of 30 species at 20 sites


#2. Diversity of each site
simpson <- diversity(dune, "simpson")
round(simpson,3)
#calculate Simpson's 1-D Index of Diversity for each site. 
# closer to 1 = greater diversity

shannon <- diversity(dune) # note that Shannon's is default
round(shannon,3) #Typically ranges from 1.5 - 3.4, higher = more diverse

# lets compare the two
par(mfrow = c(1, 2))  
hist(simpson)
hist(shannon)


#3. Calculate dissimilarity distance
# Vegdist computes dissimilarity indices. 
#We are using gower and bray-curtis which are good in
# detecting underlying ecological gradients
#They are bounded between 0 and 1, 
#  where 0 = same composition, 1 = maximally dissimilar
par(mfrow = c(1, 2))
bray = vegdist(dune, "bray") 
gower = vegdist(dune, "gower")
hist(bray, xlim = range(0.0,1.0))
hist(gower, xlim = range(0.0,1.0))


#4. Non-Metric Multidimensional Scaling (nMDS)
# The goal of NMDS is to collapse information from multiple
#  dimensions (e.g, from multiple communities, sites, etc.) 
#  into just a few, so that they can be visualized & interpreted.
# NMDS does not produce a statistical output

#create a random comunity matrix
set.seed(2) # random no. generator / way to specify seeds
community_matrix=matrix(
   sample(1:100,300,replace=T),nrow=10, 
   dimnames=list(paste("community",1:10,sep=""),
                 paste("sp",1:30,sep="")))
community_matrix[1:5,1:10]

#run nMDS
example_NMDS=metaMDS(community_matrix, # Our matrix
             k=2) # The number of reduced dimensions. 
                  #  Increase if high stress is problem.
#A good rule of thumb:
#  stress > 0.05 provides an excellent representation 
#               in reduced dimensions,
# > 0.1 is great,
# >0.2 is good/ok,
# and stress > 0.3 provides a poor representation

par(mfrow = c(1, 1))
plot(example_NMDS)

#add text to make sense of plot:
  #Ordination plot function especially for congested plots
ordiplot(example_NMDS,type="n")

  #The function adds text or points to ordination plots
orditorp(example_NMDS,display="species",col="red",air=0.01)

orditorp(example_NMDS,display="sites",cex=1.1,air=0.01)


#5. Multivariate analysis of variance for statistical testing
#We will use adonis(): 
#   A multivariate analysis of variance using distance matrices
# Adonis and uses permutation
# tests to inspect the significances of those partitions  

#Other methods implemented in vegan: 
# multiresponse permutation procedure (mrpp)
# analysis of similarities (anosim)

# create dummy data (3 sets of 30 sites, 10 species)
set.seed(666)
num<-30
disp.a<-5
sites.a<-data.frame(
  sp.a=rnbinom(num,mu = 4, size = disp.a),
  sp.b=rnbinom(num,mu = 6, size = disp.a),
  sp.c=rnbinom(num,mu = 5, size = disp.a),
  sp.d=rnbinom(num,mu = 7, size = disp.a),
  sp.e=rnbinom(num,mu = 1, size = disp.a),
  sp.f=rnbinom(num,mu = 18, size = disp.a),
  sp.g=rnbinom(num,mu = 10, size = disp.a),
  sp.h=rnbinom(num,mu = 8, size = disp.a),
  sp.i=rnbinom(num,mu = 4, size = disp.a),
  sp.j=rnbinom(num,mu = 5, size = disp.a))

disp.b<-15
sites.b<-data.frame(
  sp.a=rnbinom(num,mu = 20, size = disp.b),
  sp.b=rnbinom(num,mu = 30, size = disp.b),
  sp.c=rnbinom(num,mu = 25, size = disp.b),
  sp.d=rnbinom(num,mu = 35, size = disp.b),
  sp.e=rnbinom(num,mu = 5, size = disp.b),
  sp.f=rnbinom(num,mu = 90, size = disp.b),
  sp.g=rnbinom(num,mu = 50, size = disp.b),
  sp.h=rnbinom(num,mu = 40, size = disp.a),
  sp.i=rnbinom(num,mu = 20, size = disp.b),
  sp.j=rnbinom(num,mu = 25, size = disp.a))


disp.c<-30
sites.c<-data.frame(
  sp.a=rnbinom(num,mu = 40, size = disp.c),
  sp.b=rnbinom(num,mu = 60, size = disp.a),
  sp.c=rnbinom(num,mu = 50, size = disp.b),
  sp.d=rnbinom(num,mu = 70, size = disp.c),
  sp.e=rnbinom(num,mu = 10, size = disp.c),
  sp.f=rnbinom(num,mu = 180, size = disp.c),
  sp.g=rnbinom(num,mu = 100, size = disp.c),
  sp.h=rnbinom(num,mu = 80, size = disp.a),
  sp.i=rnbinom(num,mu = 40, size = disp.c),
  sp.j=rnbinom(num,mu = 50, size = disp.c))

#combine data
all.sites<-rbind(sites.a,sites.b,sites.c)

#treatments
trt<-rep(c("Control","High","Low"),each=nrow(sites.a))

#run nMDS
library(ggplot2)
all.mds <- metaMDS(all.sites)  #using all the defaults

data.scores <- as.data.frame(scores(all.mds))  
data.scores$site <- rownames(data.scores)  
data.scores$grp<-trt

ggplot(data=data.scores) + 
  stat_ellipse(aes(x=NMDS1,y=NMDS2,colour=trt),level = 0.50) +
  geom_point(aes(x=NMDS1,y=NMDS2,shape=trt,colour=trt),size=4)

# there seems to be some differences but let's run the stats

#run adonis
adon.results<-adonis(all.sites ~ trt, method="bray",perm=999)
print(adon.results)

#So there are S differences between control, low and high 
#....(whatever low and high are)


#Let's run this thru anosim()
all.sites.dist <- vegdist(all.sites) #dissimilarity matrix
anos.results<-anosim(all.sites.dist,trt)
print(anos.results)
#Interpretation: 
#ANOSIM tests whether we can reject the null hypothesis that
# the similarity between groups is greater than or equal
# to the similarity within the groups
# ANOSIM generates a value of R which lies between -1 and 1, 
# R=0 representing the null hypothesis (no difference
#     among treatments).
# R>0 suggest more similarity within treatment 
# R<0 suggest more similarity between sites than within sites 
#     and may raise the possibility of wrong assignment of 
#     samples to sites.
# The closer R is to 1, the more the samples within a site
#  are similar to each other and dissimilar to samples in 
#  other sites
plot(anos.results)

```