---
title: "Lesson 9. Other bits and pieces"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "C:/Matias/Cursos/2019_Mexico/htmls") })

output: 
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: yes
    theme: united
---

<style type="text/css">
h1.title {
  font-size: 30px;
    color: Maroon;
  text-align: center;
          }
h3.subtitle {
  font-size: 22px;
    color: Maroon;
  text-align: center;
            }
h4.author { 
    font-size: 24px;
      color: FireBrick;
  text-align: center;
          }
h4.date { 
  font-size: 18px;
  text-align: center;
        }
h1{
  font-size: 22px;
  color: DarkBlue;
  }
h2{
  font-size: 20px;
  color: Blue;
  }
h3{
  font-size: 18px;
  color: SteelBlue;
  }
body{
    font-family: Helvetica;
    font-size: 14pt;
    }
code.r{
  font-size: 16pt;
      }
pre {
  font-size: 16pt;
}
</style>


```{r globaloptions, include=FALSE}
knitr::opts_chunk$set(fig.width = 6,fig.height = 6,error=TRUE,
                        echo = TRUE, warning=FALSE,message=FALSE)
```
# Lesson goals
We're going to touch on other useful concepts and applications:

- Version control and Rmarkdown
- Make code more efficient


# Version control
source:
https://github.com/adw96/biostat561/blob/master/lecture1/lecture1.Rmd

What problems can you see with the following approach to version control?

- papersims-v1.R
- papersims-v2.R
- papersims-thea-comments.R
- papersims-hellfire.R
- papersims-v5.R
- papersims-final.R

1. How many versions until this becomes intractable?
2. Exponential file number growth with multiple collaborators!

...could use Dropbox ...but there's a better way..

# Git

- git is an open source version control system (VCS):
    - Track changes to code and documents. *What* changes by *who* and *when*?
    - Share code and collaborate
- GitHub is a website that uses git's VCS:
    - Collaborate with others effectively
    - Distribute code
    - Solicit improvements (*pull* requests)
    - Track issues & feature requests
    - (Build your coding portfolio!)

# Git with R

git & GitHub are popular with R developers

- Integration with RStudio
- Easy distribution of packages
    - Circumvents CRAN moderators; for better or worse
- Always get the latest features (`devtools`; `install_github`)


# Getting started with git

- Download/update: [git-scm.com/downloads](git-scm.com/downloads)
- Intro: [guides.github.com/activities/hello-world](guides.github.com/activities/hello-world)
- Do this with Homework 1 and a blank pdf, not hello-world
- Questions? Error messages? The internet is a great resource!
- *git is a great habit to get into early! Start now!*


The standard workflow for uploading a new file or updating an old one (e.g., for your homework submission):

```{r, eval = FALSE}
git pull
git commit -a -m 'question 2 part b response'
git push
```


<br><br>

# Rmarkdown

install.packages("rmarkdown")

source:
https://rmarkdown.rstudio.com

cheat sheet
https://rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf

reference guide
https://rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf

Rmarkdown is used to 

- run analysis,
-  generate outputs
-  combine those outputs into documents (pdfs, htmls, word, ppt, etc)

Each .Rmd file has three types of components:

- Meta data (YAML) that guides how to create the document
- chunks (that execute R code)
- text

**You can write your entire thesis this way (can even add your Reference list...)**
<br><br>

# Start a new R Markdown file in R Studio
1. Open R Studio, then go to File → New file → R Markdown. (Choose a title and author, but leave the rest as it is, and press the OK button)
2. Click button 'knit' (The R markdown file you opened, already contains example code)
3. Embed some R code:
```{r}
# ```{r}
#do a summary of data frame cars
summary(cars)

#plot something
par(mfrow=c(2,1))
plot(cars)
image(volcano, col=terrain.colors(50))
#```
```
The website http://rmarkdown.rstudio.com/authoring_pandoc_markdown.html provides more information on how to include lists, pictures, etc. Here only a few things are discussed.

- header 1: # header 1
- header 2: ## header 2
- header 3: ### header 3
- header 4: #### header 4

bold text: **bold** text
italics text: *italics* text
code text: `code` text

<br>

## Student task

- Build this simple power point presentation


<br> 

# Make code more efficient 

## Package purrr
source:
https://towardsdatascience.com/functional-programming-in-r-with-purrr-469e597d0229

```{r}
library(purrr)

#1. When starting using R
employee <- c('John Doe','Peter Gynn','Jolie Hope')
salary <- c(21000, 23400, 26800)
startdate <- as.Date(c('2010-11-1','2008-3-25','2007-3-14'))

# from dataframe  take mean of salary column
employ_data <- data.frame(employee, salary, startdate)
mean(employ_data$salary)

#let's do some anovas the traditional way...
aov_mpg <- aov(mpg ~ factor(cyl), data = mtcars)
summary(aov_mpg)
aov_disp <- aov(disp ~ factor(cyll), data = mtcars)
summary(aov_disp)
aov_hp <- aov(hp ~ factor(cyl), data = mrcars)
summary(aov_hpp)
aov_wt <- aov(wt ~ factor(cyl), datas = mtcars)
summary(aov_wt)
#Ooops, I copied and pasted multiple times and made some errors...


#2. OK, we can do loops...but now we know dplyr...
library(dplyr)

# Let’s calculate the R-squared values for the linear 
# relationship between Weight and Miles per Gallon, 
#  according to the number of Cylinders

  #1. create df for 4 cylinder cars
cyl_4 <- filter(mtcars, cyl == 4)
  
  #2.  create a linear model on 4 cyl cars
lm_4 <- lm(mpg ~ wt, data = cyl_4)

  #3.  get the summ
lm_4_summary <- summary(lm_4)

  #4.  get the r.squared value
lm_4_r_squared <- lm_4_summary["r.squared"]

  #5.  check the value
lm_4_r_squared


#...or we can use pipes..
lm_4cyl_rsquared <- mtcars %>% 
  filter(cyl == 4) %>%
  lm(mpg ~ wt, data = .) %>% 
  summary() %>% 
  .$"r.squared"
lm_4cyl_rsquared


#...but...to avoid repetition, we can use purrr
# purrr allows you to map functions to data
table(mtcars$cyl)

# Generates an output from all 3 of our linear models 
#   according to number of cylinders in 5 lines of code:
mtcars %>%
  split(.$cyl) %>%
  map(~ lm(mpg ~ wt, data = .)) %>%
  map(summary) %>%
  map_dbl("r.squared")


```

## Vectorising
vectorising involves manipulating your data to avoid using loops
```{r}
#Create some data
N=1e5
dat=data.frame(var1=as.numeric(1:N))

#Create var2 which equal to the cumulative sum of var1

#1. loop approach
dat$var2.loop=NA
system.time({for(i in 2:nrow(dat)) dat$var2.loop[i]=dat$var1[i-1]+dat$var1[i]})

library(dplyr)
system.time({dat=dat%>%mutate(var2.lag=lag(var1,1)+var1)})

head(dat)
```
For N=3e5, system time for loop is **183s** and for lag() is **0s**

...so exponential increase for loops
