---
title: "Lesson 1. R basics"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "C:/Matias/Cursos/2019_Mexico/htmls") })

output: 
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: yes
    theme: united
---

<style type="text/css">
h1.title {
  font-size: 30px;
    color: Maroon;
  text-align: center;
          }
h3.subtitle {
  font-size: 22px;
    color: Maroon;
  text-align: center;
            }
h4.author { 
    font-size: 24px;
      color: FireBrick;
  text-align: center;
          }
h4.date { 
  font-size: 18px;
  text-align: center;
        }
h1{
  font-size: 22px;
  color: DarkBlue;
  }
h2{
  font-size: 20px;
  color: Blue;
  }
h3{
  font-size: 18px;
  color: SteelBlue;
  }
body{
    font-family: Helvetica;
    font-size: 14pt;
    }
code.r{
  font-size: 16pt;
      }
pre {
  font-size: 16pt;
}
</style>


```{r globaloptions, include=FALSE}
knitr::opts_chunk$set(fig.width = 12,fig.height = 8,
                        echo = TRUE, warning=FALSE,message=FALSE)
```

## Repositories
all scripts stored in

https://github.com/JuanMatiasBraccini/Git_R_course


<br>
Credits to Ainslie Denham, Alex Hesp and John Hoenig
<br><br>

# Introduction to R and Rstudio
1.  A script is a combination of **comments** and **words** which is R syntax (code)
```{r }
#calculate mean
mean(1:10)
```
2.  **Rstudio** provides an interface for using the program R and it has 4 main panes:
+ top left is where you write code (this is where you save your script)
+ below is the console where you see outputs and code that you have run
+ top right window shows info on objects 
+ bottom right window shows plot, help, etc.

<br><br>

## Using R like a calculator
```{r }
# what is 4 + 5?
#type 4+5 and press 'Run' or 'ctrl+Enter' (this is how we run code)
4 + 5

# Multiplication, let's try 21 * 6
21 * 6

# Subtraction
63 - 2

# Division
10 / 5

# A mixture of basic operations
2 * (10 + 6)
```

<br><br>

## Various ways to run code
+ To run a single line of code, position the cursor and press 'Run' or 'Ctrl + Enter'
+ To run several lines of code at once, highlight them all and press 'Run' or 'Ctrl + Enter'
+ To run whole code at once, 'Ctrl + a' and 'Run'
+ To run all code above, 'Ctrl + shift + home' and 'Run'


<br><br>


## Storing values as "variables"
```{r}
 # Store information in computer memory, so that we can use it later
FirstNumber <- 4    
SecondNumber <- 5
#note: when naming variables use names that describe what information 
#      is being stored, so that the name means something to you
Result=FirstNumber+SecondNumber 
Result   #the new variable (where the outcome is stored)

#can use '<-' or '=' but note that 
#                 the operator <- can be used anywhere, whereas 
#                 the operator = is only allowed at the top level
#                 x <- y <- 5
#                 x = y = 5
#                 x = y <- 5
#                 x <- y = 5


# In the top right window (environment window), you should see under 
# 'Values' our variable names
# So the computer now knows that our first variable, FirstNumber,
# is equal to 4, and our second variable, SecondNumber, is 
# equal to 5

#note that R is case-sensitive, type 'firstnumber'

#if not interested in keeping a variable in the memory
rm(FirstNumber)
# so it's now gone from the environment window.

#if we need to clear the entire memory
rm(list=ls()) #if we put this at the beginning of a script, we start
#               with no info being stored in memory

# Text variables (called 'strings')
CalendarMonth <- "Jan"   #note how we need to add " " around the text
CalendarMonth
```

<br><br>

## Storing several values as a vector
```{r}
# Often we want to store, manipulate, etc., multiple numbers at once
# For example, we can store several numbers in a "vector", which is 
#  collection of values (numeric or characters, i.e. strings).
NumberVector1 <- c(1,2,3,4,5)
NumberVector1    #look at the top right pane...

#vector calculations
NumberVector2 <- c(6,7,8,9,10)
NumberVector1 + NumberVector2  #(element wise)
NumberVector1 * NumberVector2

#joining vectors
NumberVector3 <- c(NumberVector1,NumberVector2)

# what about text
TextVector1 <- c("Jan","Feb","Mar")
TextVector2 <- c("Apr","May","Jun")
TextVector3=c(TextVector1,TextVector2)  #type TextVector1+TextVector2

#Note: the data in a vector must be only one type or mode (numeric,
#      character or logical). You can’t mix them in the same vector
```

<br><br>

## Types of vectors
* The “type” of a vector refers to the elements within the vector.
* There are six “types” of vectors, for example:
  * numeric:
      * “integer” (e.g., 5)
      * “double” (e.g., 5.5)
  * character (e.g., “surfing”)
  * logical (e.g., TRUE , FALSE )

You can use typeof() function to examine vector type

All vectors can be “named” (i.e., name individual elements within vector)
```{r}
y<-c(student1="this course sucks",student2="this is awesome")
y
```

<br><br>

## Student task
1. Create a vector called v1 with three elements, where all the elements are
numbers. Then print the values.
2. Create a vector called v2 with four elements, where all the elements are
characters (i.e., enclosed in ' ' or " " quotes). Then print the
values.
3. Create a vector called v3 with five elements, where some elements are
numeric and some elements are characters. Then print the values.


<br><br>

## Solution to student task
```{r}
# create vector v1 with three elements all the elements are numbers
v1 <- c(1, 2, 3)
# print value
v1

# create vector v2 with four elements all the elements are characters
v2 <- c("a", "b", "c", "d")
# print value
v2 

v3 <- c(1, 2, 3, "a", "b")
# create vector v3 with five element some elements are numeric
#   and some elements are characters
# print value
v3
typeof(v3)
```


<br><br>

## Specifying number sequences
```{r}
# how can we create the numbers 1,2,.., 10, without writing them all?
seq(from=1,to=10,by=1)
# this is the same as
seq(1,10,1)  # seq is an inbuilt function (it has arguments); 
# If you don't know the function/arguments: ask Google, type ?seq

#Let's store these numbers
NumberVector4 <- seq(1,10,1)
NumberVector4

# Intro to vectorisation (i.e. doing many calculations with a single
#   command)using a fisheries example
#   The annual survival rate, S, is related to the total
#     instantaneous mortality rate, Z, by S = exp(-Z).
#   It follows that Z is related to S by Z = - log(S)
#     where log means natural logarithm.
# Suppose we want to convert Z = .2 into a survival rate. We can do:
Z = .2
S = exp(-Z)
S

# Suppose we want to convert several values of Z (e.g. Z at age)?
Z = seq(from=.05,to=1.5,by=.05)
S = exp(-Z)
# There you have it: 30 values of Z and the corresponding values of S. 
cbind(Z,S)


#Exercise: create a sequence of numbers starting at 0, finishing 
#          at 100, and incrementing by 2

#Let's reverse the sequence
seq(10,1,-1)
rev(seq(1,10,1)) 
#note: there are multiple ways of getting to the same solution....

#Exercise: create a sequence of numbers increasing from 0 to 20 by 2
#           and then decreasing back to 0 by 2 without repeating 20 

```

<br><br>

## Basic summary statistics
```{r}
NumberSequence <- c(seq(0,20,2),seq(18,0,-2))
length(NumberSequence)        #how namy numbers in our sequence?
max(NumberSequence)           #what is the maximum value?
min(NumberSequence)           #what is the minimun value?
range(NumberSequence)         #what is the range?
sum(NumberSequence)           #what do all the numbers add up to?
mean(NumberSequence)          #what is the mean?
sd(NumberSequence)            #what is the standard deviation?
median(NumberSequence)        #what is the 50th percentile (median) 
#...type percentile(NumberSequence)
quantile(NumberSequence,0.75) #what is the 75th percentile
summary(NumberSequence)       #summary

```


<br><br>


## Vectors of repeated numbers
```{r}
rep(1,5)
rep(2,3)
c(rep(1,5),rep(2,3))  #combined results
rep(1:2, each=2)     #repeat each times

```

<br><br>


## Basic functions
* Assignment
  * <- and =
* Creating data structures
  * seq( )
  * c( )
  * rep( )
  
* Built-in functions
  * **abs( )**       absolute value
  * **as.matrix()**  coerces a data structure to be a matrix
  * **cat( )**       print a message
  * **dim( )**       dimensions
  * **is.na( )**     logical: returns TRUE/FALSE 
  * **length( )**    length of a vector (or list)
  * **matrix( )**    create a matrix
  * **max( )**       maximum value
  * **mean( )**      mean value
  * **min( )**       minimum value
  * **plot( )**      creates a plot of y vs x
  * **prod( )**      gives the product of the numbers
  * **range( )**     minimum and maximum values
  * **sum( )**       sum of elements
  * **var( )**       calculates the sample variance

## Cheat sheet
https://www.rstudio.com/wp-content/uploads/2016/10/r-cheat-sheet-3.pdf



<br><br>


## Accessing elements within a vector (subsetting)
```{r}
NumberSequence[1] # the first number of that sequence
NumberSequence[5] # the fifth number of that sequence
NumberSequence[1:5] # the first five numbers of that sequence
NumberSequence[c(1,5)] # the first and fifth numbers of that sequence

#what if we want to find a particular number in the vector?
id=which(NumberSequence==4) #the == indicates a logical operation
id
NumberSequence[id]

#find the values NOT EQUAL to 4
id=which(!NumberSequence==4) 
NumberSequence[id]

#which ones had a value less than or equal to 10?
id=which(NumberSequence<=10)
NumberSequence[id]

#which ones had a value more than 10?
id=which(NumberSequence>10)
NumberSequence[id]

# which ones are less/equal to 4 OR  greater/equal to 18?
id=which(NumberSequence<=4 | NumberSequence>=18)
NumberSequence[id]

#which ones are less/equal to 10 AND greater/equal to 4?
id=which(NumberSequence<=10 & NumberSequence>=4)
NumberSequence[id]

#logical operations for character vectors
which(TextVector1=="Feb")
which(TextVector1%in%c("Feb","Mar"))
id=which(TextVector1%in%c("Feb","Mar"))
TextVector1[id]
TextVector1[which(TextVector1%in%c("Feb","Mar"))]
which(month.abb=="Jun")     # vector already defined in R
which(month.name=="April")  # vector already defined in R

```

<br><br>

## Wrap up exercise
1. Create a vector containing the numbers 1 to 10 
2. Create a second vector containing the numbers 1 to 5
3. Join them together
4. Add 5 to each number, then multply the resultant numbers by 6
5. Calculate the minimum, maximum and mean of the resultant set of numbers
6. Work out which of the numbers are greater then 60 
7. Subtract 10 from only those numbers that are greater than 60, and add them all up!


<br><br>

# Matrices
All elements have same mode in matrices (e.g. all numeric)
```{r}
Mat=matrix(c(1,2,3,5),nrow=2,byrow=T)
Mat

#Be careful with matrix operations
vec=c(0.5,2)
Mat*vec  #elementwise
Mat%*%vec #matrix multiplication, 
#i.e. firt row= 1*0.5 + 2*2
#     second row= 1*0.5 + 5*2

# Overview of matrix operations in R:
#  Syntax          Operation
#  *, +, -, /      elementwise multiplication, addition, etc
#  %*%             matrix multiplication
#  t(A)            transpose of A
#  crossprod(A,B)  equivalent to t(A)%*%B
#  solve(A,b)      solve the linear system of equations Ax = b
#  solve(A)        find the inverse of the matrix A
#  chol(A)         Cholesky decomposition of A
#  qr(A)           QR decomposiiont of A
#  svd(A)          singular value decomposition of A
#  eigen(A)        eigen decomposition of A
#  kronecker(A,B)  Kronecker product of A & B
#  diag(b)         create a diagonal matrix from the vector b
#  diag(A)         extract the diagonal elements of A

b = c(1,2,3) # create a vector
diag(b)      # create a diagonal matrix

A = matrix(1:9,nrow=3)
A
diag(A)      # extract the diagonal elements of matrix A
t(A)         # take the transpose of A

# an important note about subsetting matrices: 
# if you extract one column or one row of a matrix
# the default is for R to make it into a vector.
A
A[,1]
# You can keep the result as a matrix by using the drop=F
# option as follows:
A[,1,drop=F]
# This is important because if you want to do matrix 
# multiplication you might need the vector to be in
# the right matrix form.

# solve the system of equations:
# 2 = 3y + 2z
# 1 = 2y + 4z
(b = c(2,1))
(A = matrix(c(3,2,2,4),nrow=2,byrow=T))
solve(A,b)
# we see that y = .75 and z = -.125
```


<br><br>


# Lists

* What is a list?
    * Like vectors, a list is an object that contains elements
    * Unlike vectors, these elements can differ in class (numeric, character, 
      data frames, other lists) and length
    * Ideal of storing different kind of data types
  
```{r}
mylist=list(a=1:5,
            b=rep('my first list',2),
            data1=matrix(1:10,ncol=2),
            sublist=list(x=1,y=1:10))
typeof(mylist)
length(mylist)
names(mylist)
mylist
#rather than printing
str(mylist)

#Accesing list elements
mylist[[1]]
mylist[[4]]
mylist[[4]][[1]]
mylist$sublist$x
            
```