---
title: "Lesson 6. Plotting and Mapping"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "C:/Matias/Cursos/2019_Mexico/htmls") })

output: 
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: yes
    theme: united
---

<style type="text/css">
h1.title {
  font-size: 30px;
    color: Maroon;
  text-align: center;
          }
h3.subtitle {
  font-size: 22px;
    color: Maroon;
  text-align: center;
            }
h4.author { 
    font-size: 24px;
      color: FireBrick;
  text-align: center;
          }
h4.date { 
  font-size: 18px;
  text-align: center;
        }
h1{
  font-size: 22px;
  color: DarkBlue;
  }
h2{
  font-size: 20px;
  color: Blue;
  }
h3{
  font-size: 18px;
  color: SteelBlue;
  }
body{
    font-family: Helvetica;
    font-size: 14pt;
    }
code.r{
  font-size: 16pt;
      }
pre {
  font-size: 16pt;
}
</style>


```{r globaloptions, include=FALSE}
knitr::opts_chunk$set(fig.width = 6,fig.height = 6,error=TRUE,
                        echo = TRUE, warning=FALSE,message=FALSE)
```

# Lesson goals
We're going to expand on doing plots and maps.



full list of packages necessary for this series of tutorials can be installed with:

install.packages(c("cowplot", "googleway", "ggplot2", "ggrepel", "maps","rgeos","grid","ggspatial", "libwgeom", "sf", "rnaturalearth", "rnaturalearthdata","rsvg","magick","PBSmapping","tmap","spData",
"mapview", "leaflet")

# Base Plotting
Credits to John Hoenig
<br>
## 1. par()
```{r}

#type ?par
plot(1:10)
par(bg=3,cex.axis=2,cex.lab=3,
    col.axis="steelblue",fg="white",
    font.axis=2,las=3)
plot(1:10)

dev.off() #rest par to default

```

## 2. Margins
```{r}
# There are two kinds of margins in a plot:
#    plot margins - where you put your axis labels for e.g.
#    outer margins - where you can put text, especially when
#                    you have several plots on a page.
#                    DEFAULT is to have outer margins = 0

# create some sample data to plot:
x1 = 2:5
y1 = c(8,5,4,1)
x2 = 2:5
y2 = c(2,3,5,7.7)

# set the outer (figure) margin & the inner (plot) margin
par(oma=c(4,4,4,4),mar=c(4,4,4,4))  
plot(x1,y1,typ="l", ylab="This is y",
     xlab="This is x") 

# red box around 
box(which="plot",lty=1,col=2)   

# green box around the figure margin
box(which="figure",lty=2,col=3) 

# add some text in the inner margin of side 1
mtext("plot margin 1",side=1,line=2,col=2)  

# add text in the outer margin on side 1
mtext("outer (figure) margin 1",side=1,line=2,outer=T,col=3)  

# text on side 4 illustr. line option
mtext("plot margin 4 in line 1",side=4,col=2,line=1)
mtext("plot margin 4 in line 2", side = 4,col=2,line=2)

# same, for outer margin
mtext("outer margin 4 inline 1",side=4,col=3,line=1,outer=T)
mtext("outer margin 4 in line 2",side=4,col=3,line=2,outer=T)

```


## 3. Margins for multiplots
```{r}
par(mfrow=c(2,2),oma=c(4,4,4,4),mar=c(4,4,4,4))
plot(x1,y1,typ="l")    

# box around plot
box(which="plot",lty=1,col=2)    
box(which="figure",lty=2,col=3) 
# NOTE where the box goes!

mtext("plot margin 1",side=1,line=2,col=2)  
plot(x2,y2,typ="l")              
box(which="plot",lty=1,col=2) 
box(which="figure",lty=2,col=3)
mtext("plot margin 1",side=1,line=2,col=2)  
# add text to the outer margin: NOTE where the text goes!

mtext("outer margin 2",side=2,line=2,col=3,outer=T)

# make our 3rd graph, but change the margins
par(mar=c(4,4,0,0))     
plot(x1,y1,typ="l")
box(which="plot", col=2,lty=1)

# let's add the box around the outside now
box(which="figure",col=3,lty=2)

# let's do the last graph
par(mar=c(4,4,4,4))
plot(x2,y2,typ="l")
mtext("inner margin, line=0",line=0,col=2,side=1)
mtext("inner margin, line=2",line=2,col=2,side=1)
mtext("inner margin, line=4",line=4,col=2,side=1)

# add some text to the outer margins
mtext("outer margin 1",line=2,col=3,side=1,outer=T)
mtext("outer margin 3",line=2,col=3,side=3,outer=T)
mtext("outer margin 4",line=2,col=3,side=4,outer=T)
```
<br> 

## Student task 1.
- Create five vectors of X (X1 to X5) and of Y (Y1 to Y5)
- Plot each X-Y pair (5 plots in total) in one single page.
Plot each graph with a different point color and type (see 'pch'),
also use different colors for the axis. Finally remove the X and Y label of each graph and instead have one single overall X and Y axis label (tip, use mtext(...outer=T)). Make sure all text (axis labels and titles) are vertical (tip: control thru par())

## 4. Palette (one way to control colors)
```{r}
palette('default')       # obtain the current palette
plot(1:6,col=1:6,pch=19,cex=3)

palette(rev(rainbow(6)))     # six color rainbow
plot(1:6,col=1:6,pch=19,cex=3)

(palette(gray(seq(0,.9,len = 25)))) # gray scales
plot(1:25,col=1:25,pch=19,cex=3)


par(mfcol=c(1,1))
# Here is a little program to show you what
# some color pallettes look like.
# Don't worry about trying to understand 
# the program - just look at the output
# to see some of your options

demo.pal <-function(n, border = if (n<32) "light gray" else NA,
           main = paste("color palettes;  n=",n),
           ch.col = c("rainbow(n, start=.7, end=.1)", "heat.colors(n)",
                      "terrain.colors(n)", "topo.colors(n)",
                      "cm.colors(n)"))
  {
    nt <- length(ch.col)
    i <- 1:n; j <- n / nt; d <- j/6; dy <- 2*d
    plot(i,i+d, type="n", yaxt="n", ylab="", main=main)
    for (k in 1:nt) {
      rect(i-.5, (k-1)*j+ dy, i+.4, k*j,
           col = eval(parse(text=ch.col[k])), border = border)
      text(2*j,  k * j +dy/4, ch.col[k])
    }
  }
demo.pal(20)
```


## 5. 3d graphs
```{r}
# There are some other graphical functions of interest in the
#   lattice and MASS packages. 
#  Here's a table summarizing availability of functions in
#   the graphics package (which comes with R)
#   and the MASS & lattice packages, which you have 
#   to load, e.g., using library(MASS)

# METHOD                  IN (graphics)         IN (package)
#
# 3D scatterplot                                cloud (lattice)
# matrix of scatterplots  pairs                 splom (lattice)
# bivariate density surf. persp                 wireframe (lattice)
# contour plot            contour, image       contourplot (lattice)
#                         contourLines          contour (MASS)
#                         filled.contour        levelplot (lattice)
# parallel coord. plot                          parallel (lattice)
#                                               paracoord (MASS)
# star plot               stars
# segment plot            stars
# interactive 3D graphics                       (rggobi), (rgl)

# For example, let's do a contour plot
x <- -6:16
op <- par(mfrow = c(2, 2),mar=c(2,2,1,1))
contour(outer(x, x), method = "edge", vfont = c("sans serif", "plain"))
z <- outer(x, sqrt(abs(x)), FUN = "/")
image(x, x, z)
contour(x, x, z, col = "pink", add = TRUE, method = "edge",
        vfont = c("sans serif", "plain"))
contour(x, x, z, ylim = c(1, 6), method = "simple", labcex = 1,
        xlab = quote(x[1]), ylab = quote(x[2]))
contour(x, x, z, ylim = c(-6, 6), nlev = 20, lty = 2, method = "simple",
        main = "20 levels; \"simple\" labelling method")

```


## 6. Expressions
```{r}
# Here we will use the paste() and expression() functions to print
#  text and equations that incorporate special 
# symbols and mathematical notation.

#?expression    # the help page is not very useful...

# expression() creates an expression that can be printed.
# It does it by translating code for equations and 
#  special symbols into plotting instructions and 
#  pasting those translated equations together with 
#  any character strings you specify.

# try executing these commands

plot(0,0,col=0,xlim=c(0,20),ylim=c(0,20))  # set up a plot

#add some expressions
text(10,20,paste("likelihood = ",pi^3*(1-pi)^7),cex=1.5)   
text(10,17,paste("likelihood = ","pi^3*(1-pi)^7"),cex=1.5)
text(10,14,expression(paste("likelihood = ",pi^3*(1-pi)^7)),cex=1.5)
text(10,10,expression(paste("area = ",pi * r^2)))
text(10,7,expression(paste("f(x) = ",1/sqrt(2*pi * sigma)*exp(- 1/(2*sigma^2)*(x - mu)^2))))
text(10,5,expression(paste("f(x) = ",1/sqrt(2*pi * sigma)*e^- 1/(2*sigma^2)*(x - mu)^2)))
text(10,3,expression(paste("f(x) = ",1/sqrt(2*pi * sigma)*exp(- (x - mu)^2/(2*sigma^2)))))

# instruction manual for typsetting equations:
# demo(plotmath)    

```
<br>

## Student task 2

Plot X (X=1:10) and Y (Y=X*3), each point as a filled circle
of progressively increasing shades of grey (tip, modify the palette but then change it back to default)
Add this expression to the X axis:
$$\sum_{i=1}^{n} X_i^y$$

<br>

## 7. Multiplots in one page
```{r}
n=10
x1 <- runif(n,1,10)
y1 <- x1*2*runif(n,.8,1.2)
x2 <- x1*2
y2 <- x2*2*runif(n,.8,1.2)
x3 <- x1*3
y3 <- x3*2*runif(n,.8,1.2)
x4 <- x1*4
y4 <- x4*2*runif(n,.8,1.2)

#  par() 
par(mfrow=c(4,2))
plot(x1,y1)
plot(y1,x1)
plot(x2,y2)
plot(y2,x2)
plot(x3,y3)
plot(y3,x3)
plot(x4,y4)
plot(y4,x4)

#  layout()
M=matrix(c(2,0,1,3),2,2,byrow=T)  
M
# this shows we want to draw the first plot at the
# bottom left, the second at top left, the
# third at bottom right, and we want blank
# space at top right (indicated by 0)
layout(M)
plot(x1,y1,main="graph 1")
plot(x2,y2,main="graph 2")
plot(x3,y3,main="graph 3")

# other options: 
# split.screen( ) 
# par(fig=c(x1,x2,y1,y2))
```
<br><br>

# GGPLOT

source

http://r-statistics.co/Complete-Ggplot2-Tutorial-Part1-With-R-Code.html

http://r-statistics.co/Complete-Ggplot2-Tutorial-Part2-Customizing-Theme-With-R-Code.html

http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html

```{r}
library(ggplot2)

#Basics of ggplot syntax:

#1. load some data
data("midwest", package = "ggplot2") 


#2.  Initiate ggplot
ggplot(midwest, aes(x=area, y=poptotal))
# area and poptotal are columns in 'midwest'
head(midwest,2)

# what happened?
# A blank ggplot is drawn. Even though the x and y are specified,
#  there are no points or lines in it. 
# This is because, ggplot doesn’t assume that you meant a
#  scatterplot or a line chart to be drawn.
# Basically, we have to build the ggplot adding layer by layer
#The aes() function is used to specify the X and Y axes


#3. Let's make a simple scaterplot
ggplot(midwest, aes(x=area, y=poptotal)) +
  geom_point()


#4. Let's add a smoothing line
g <- ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point() + 
  geom_smooth(method="lm")  # set se=FALSE to turnoff confidence bands
#note that I saved the ggplot to an object ("g") and then plotted 'g'

plot(g)



#5. Adjust X and Y by setting the axes' limits
g + xlim(c(0, 0.1)) + ylim(c(0, 1000000))
#note how I recall 'g' and build on top...


#6. Adjust X and Y by zooming in 
# (this does not delete points outside the X and Y range)
g1 <- g + coord_cartesian(xlim=c(0,0.1), ylim=c(0, 1000000))  
plot(g1)

#Since all points were considered, the line of best fit did 
# not change compare to previous graph....


#7. Change title and axis lables
g1 + 
  labs(title="Area Vs Population", subtitle="From midwest dataset", 
       y="Population", x="Area", caption="Midwest Demographics")
# or
g1 + 
  ggtitle("Area Vs Population", subtitle="From midwest dataset") +
  xlab("Area") +
  ylab("Population")


#8. Change color and size of points
ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(col="steelblue", size=3) +   # Set point color and size
  geom_smooth(method="lm", col="firebrick") +  # change the color of line
  coord_cartesian(xlim=c(0, 0.1), ylim=c(0, 1000000)) + 
  labs(title="Area Vs Population", 
       subtitle="From midwest dataset", 
       y="Population", 
       x="Area", 
       caption="Midwest Demographics")
    

#9. Change color to reflect categories in another column
gg <- ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(aes(col=state), size=3) +  # Set color to vary based 
                                        # on state categories.
  geom_smooth(method="lm", col="firebrick", size=2) + 
  coord_cartesian(xlim=c(0, 0.1), ylim=c(0, 1000000)) + 
  labs(title="Area Vs Population", 
       subtitle="From midwest dataset", 
       y="Population", 
       x="Area", 
       caption="Midwest Demographics")
plot(gg)
#Now each point is colored based on the state it belongs because 
# of aes(col=state).

# Not just color, but size, shape, stroke (thickness of boundary) 
#  and fill (fill color) can be used to discriminate groupings.

#remove legend if don't want it
gg + theme(legend.position="None")  # remove legend

#change color of palette
gg + scale_colour_brewer(palette = "Set1")  # change color palette

#you can check the palette options here
library(RColorBrewer)
head(brewer.pal.info, 10)


#9. Change axis text and ticks location
gg + scale_x_continuous(breaks=seq(0, 0.1, 0.01)) #change tick location
gg + scale_x_continuous(breaks=seq(0, 0.1, 0.01),
                 labels = letters[1:11])  #change tick location and name

#10. Reverse the X Axis Scale
gg + scale_x_reverse()


#11. Change theme 
gg + theme_bw() + labs(subtitle="BW Theme")
gg + theme_classic() + labs(subtitle="Classic Theme")
gg + theme_dark() + labs(subtitle="Dark Theme")


#12. Add plot components
gg <- ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(aes(col=state, size=popdensity)) + #change points color & size
  geom_smooth(method="loess", se=F) +   #add smoothing spline
  xlim(c(0, 0.1)) +
  ylim(c(0, 500000)) + 
  labs(title="Area Vs Population", y="Population",
       x="Area", caption="Source: midwest")
plot(gg)


#13. Modify theme components
gg + theme(plot.title=element_text(size=20, 
                                   face="bold", 
                                   family="American Typewriter",
                                   color="tomato",
                                   hjust=0.5,
                                   lineheight=1.2),  # title
           plot.subtitle=element_text(size=15, 
                                      family="American Typewriter",
                                      face="bold",
                                      hjust=0.5),  # subtitle
           plot.caption=element_text(size=15),  # caption
           axis.title.x=element_text(vjust=10,  
                                     size=15),  # X axis title
           axis.title.y=element_text(size=15),  # Y axis title
           axis.text.x=element_text(size=10, 
                                    angle = 30,
                                    vjust=.5),  # X axis text
           axis.text.y=element_text(size=10))  # Y axis text


#14. Modify legend and add text
midwest_sub <- midwest[midwest$poptotal > 300000, ]
midwest_sub$large_county <- ifelse(midwest_sub$poptotal > 300000, 
                              midwest_sub$county, "")

gg + geom_text(aes(label=large_county), size=2, data=midwest_sub) +
  labs(subtitle="With ggplot2::geom_text") +
  guides(color=guide_legend("State"), size=guide_legend("Density"))  

#move text so we can actually see it and remove legends
library(ggrepel)  #we need this library to move the test
gg + geom_label_repel(aes(label=large_county), size=2, 
                      data=midwest_sub) +
     labs(subtitle="With ggrepel::geom_label_repel") +
     theme(legend.position = "None")


#15. Multiple plots in one page
data(mpg, package="ggplot2")  # load data

g <- ggplot(mpg, aes(x=displ, y=hwy)) + 
  geom_point() + 
  labs(title="hwy vs displ", caption = "Source: mpg") +
  geom_smooth(method="lm", se=FALSE) + 
  theme_bw()  # apply bw theme
  
  #single plot
plot(g)

  #multiplot with the same scale for each plot
g + facet_wrap( ~ class, nrow=3) + 
  labs(title="hwy vs displ", caption = "Source: mpg", 
       subtitle="Ggplot2 - Faceting - 
                Multiple plots in one figure")  

# multiplot with free scales
g + facet_wrap( ~ class, scales = "free") +
  labs(title="hwy vs displ", caption = "Source: mpg",
       subtitle="Ggplot2 - Faceting - Multiple plots 
                  in one figure with free scales")  


#16. Add an image
library(grid)
library(magick)

  #import tiger image
tiger <- image_read_svg('http://jeroen.github.io/images/tiger.svg', 
                        width = 350)

  #convert image to raster
g_pic <- rasterGrob(tiger, interpolate=TRUE)

  #add image to graph
g+ annotation_custom(g_pic , xmin=5, xmax=7, ymin=30, ymax=45)
ggsave("C:/Matias/Cursos/2019_Mexico/Plots_Maps/tiger.png", width = 6,
       height = 6, dpi = "screen")

dev.off() 
```
<br>

## Student task 3
- 1. Using the data frame mpg, do a red histogram of cty with 50 bins
- 2. do a boxplot of hwy for different levels of drv
- 3. do a scatterplot of displ vs hwy and display (show points as orange) 
- 4. Now increase point size to 3 and show point color
  reflecting the number of cylinders (var cyl)
- 5. Change X axis label to "DISPL" and add main title "DISPL vs hwy"  and subtitle "some stuff"
- 6. do a scatterplot of displ vs hwy and display but
  showing each value of cyl in a different plot
  AND  all plots in the same page


<br>

# PBS mapping
PBS stands for Pacific Biological Station (PBS) in Nanaimo, British Columbia, Canada
...many packages created there...

I learnt to do maps with this package >10 years ago and I'm sticking to it...still haven't come across something I couldn't do.

```{r}
library(PBSmapping)
data(worldLLhigh)  #read in in shape file of the world


#plot the world
tiff(file="C:/Matias/Cursos/2019_Mexico/Plots_Maps/PBS_map1_The world.tiff",
     width=2400,height=1600,units="px",res=300,compression="lzw")
plotMap(worldLLhigh,main="this is the world",col="orange")
arrows(100,-50,116.45,-32,lwd=3,col="black",length=.1)
text(100,-52,"I live are here",col="black")
dev.off()

```

```{r}
#plot Australia and ad some color
OZ.lat=c(-44.5,-11) #lat and long ranges of interest
OZ.long=c(113,155)
plotMap(worldLLhigh, xlim=OZ.long,ylim=OZ.lat,
        col=rainbow(10))

# Add some text and locations
Perth=c(115.866,-31.95)

plotMap(worldLLhigh, xlim=OZ.long,ylim=OZ.lat,
        col="orange")
points(Perth[1],Perth[2],pch=19,cex=1.25)
text(Perth[1],Perth[2],"Perth",pos=4)

#change axis ticks
plotMap(worldLLhigh, xlim=OZ.long,ylim=OZ.lat,
col="orange",tck = 0.05, tckMinor = 0.015)

#add nicer axis and a box
par(mar=c(3,2,.1,.1),oma=c(2,2,1,1),las=1)
plotMap(worldLLhigh, xlim=OZ.long,ylim=OZ.lat,
col="orange",xlab="",ylab="",axes=F)
axis(side=1,at = seq(OZ.long[1],OZ.long[2],1), 
     labels = F,tcl=.25)
axis(side=1,at = seq(OZ.long[1],OZ.long[2],2), 
     labels = seq(OZ.long[1],OZ.long[2],2),tcl=-.75)
axis(side=2,at = seq(OZ.lat[1],OZ.lat[2],1), 
     labels = F,tcl=.25)
axis(side=2,at = seq(OZ.lat[1],OZ.lat[2],2), 
     labels = -round(seq(OZ.lat[1],OZ.lat[2],2)),tcl=-.75)
mtext(expression(paste("Latitude (",degree,"S)",sep="")),
      side=2,line=3,las=3,cex=1.1)
mtext(expression(paste("Longitude (",degree,"E)",sep="")),
        side=1,line=3,cex=1.1)
 
box()


#add a polygon
dev.off()
South.WA.lat=c(-36,-25); South.WA.long=c(112,129)
S.WA.long=c(South.WA.long[2], South.WA.long[2], 
            South.WA.long[1], South.WA.long[1])
S.WA.lat=c(South.WA.lat[2], South.WA.lat[1], 
           South.WA.lat[1], South.WA.lat[2])

plotMap(worldLLhigh, xlim=OZ.long,ylim=OZ.lat,
        col="steelblue")
polygon(x=S.WA.long,y=S.WA.lat,lwd=1.5,
        col=rgb(.1,1,.1,alpha=.4))
  
#Add some spatial catch information
N=100
sp.ktc=data.frame(
  Long=runif(N,118,135),
  Lat=runif(N,-37,-35),
  Ktch=rnorm(N,2,.5)
    )
plotMap(worldLLhigh, xlim=c(113,140),ylim=c(-38,-25),
        col="steelblue")
with(sp.ktc,points(Long,Lat,pch=21,col="red",bg="yellow",
                   cex=2*Ktch/max(Ktch)))

#add legend
Show=quantile(sp.ktc$Ktch,probs=c(.01,.99))
legend("topright",paste(round(Show,2)),cex=2,
       bty='n',pch=21,col="red",pt.bg = "yellow",
       pt.cex=2*Show/max(sp.ktc$Ktch))

#locator()   #get data from plot  

#Export map
tiff(file="C:/Matias/Cursos/2019_Mexico/Plots_Maps/PBS_map1.tiff",
     width=2400,height=2400,units="px",res=300,compression="lzw")
plotMap(worldLLhigh, xlim=c(113,140),ylim=c(-38,-25),
        col="steelblue")
dev.off()

#what happens if we change the height and width?
tiff(file="C:/Matias/Cursos/2019_Mexico/Plots_Maps/PBS_map2.tiff",
     width=1200,height=2400,units="px",res=300,compression="lzw")
plotMap(worldLLhigh, xlim=c(113,140),ylim=c(-38,-25),
        col="steelblue")
dev.off()

jpeg(file="C:/Matias/Cursos/2019_Mexico/Plots_Maps/PBS_map1.jpeg",
     width=2000,height=2000,units="px",res=300)
plotMap(worldLLhigh, xlim=c(113,140),ylim=c(-38,-25),
        col="steelblue")
dev.off()

pdf(file="C:/Matias/Cursos/2019_Mexico/Plots_Maps/PBS_map1.pdf")
plotMap(worldLLhigh, xlim=c(113,140),ylim=c(-38,-25),
        col="steelblue")
dev.off()

```

<br>

## Student task 4
Try to recreate map 'Exercise 4'



<br><br>

# Mapping with ggplot and sf

(note: As with many things in R, there are several other packages (e.g. sp, ggmap) for mapping and GIS, choose your pick...)

source:

https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html

https://www.r-spatial.org/r/2018/10/25/ggplot2-sf-2.html


```{r}
library("ggplot2")

#set up theme
theme_set(theme_bw()) 
# this is a classic dark-on-light theme (good for maps) 
# check other themes in ?ggtheme

library("sf")

# The package rnaturalearth provides a map of 
#  countries of the entire world (the shape file) 
library("rnaturalearth")
library("rnaturalearthdata")

# Use ne_countries to pull country data and choose the scale 
# (rnaturalearthhires is necessary for scale = "large"). 
# The function can return sp classes (default) or directly sf classes, 
# as defined in the argument returnclass:                   
world <- ne_countries(scale = "medium", returnclass = "sf")
class(world)

#1st, create a base map of the world using ggplot2.
# This base map will then be extended with different map elements
ggplot(data = world) +
       geom_sf()
#remember that layers are added one at a time in ggplot... 
#  type just ggplot(data = world)


#2nd, title and change countries' color
ggplot(data = world) +
  geom_sf(color = "orange", fill = "pink") +
  xlab("Longitude") + ylab("Latitude") +
  ggtitle("World map", 
      subtitle = paste0("(", length(unique(world$name)),
                        " countries)"))

#...or maybe use color according to country population size 
ggplot(data = world) +
  geom_sf(aes(fill = pop_est)) +
  scale_fill_viridis_c(option = "plasma", trans = "sqrt")

#ggplot is using the pop_est to fill a gradient...
world$pop_est[1:5]


#3rd, change map projection
ggplot(data = world) +
  geom_sf() +
  coord_sf(crs = "+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 
           +y_0=3210000 +ellps=GRS80 +units=m +no_defs ")
#the function 'coord_sf' allows to deal with the coordinate system, 
# which includes both projection and extent of the map.


#4th, zoom in into the Gulf of Mexico
ggplot(data = world) +
  geom_sf() +
  coord_sf(xlim = c(-102.15, -74.12), 
           ylim = c(7.65, 33.97), expand = T)
# The extent of the map can also be set in coord_sf, 
#  allowing to “zoom” in the area of interest, provided by
#  limits on the x-axis (xlim), and on the y-axis (ylim)


#5th, add scale bar
library("ggspatial")  #for scale bar
ggplot(data = world) +
  geom_sf() +
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "bl",
          which_north = "true", pad_x = unit(0.75, "in"), 
          pad_y = unit(0.5, "in"),
          style = north_arrow_fancy_orienteering) +
  coord_sf(xlim = c(-102.15, -74.12), ylim = c(7.65, 33.97))


#6th, add some text (e.g. country name)
# The world data set already contains country names 
# and the coordinates of the centroid of each country 
#  (among more information)

world_points<- st_centroid(world)   #from package 'sf'
world_points <- cbind(world, 
                      st_coordinates(st_centroid(world$geometry)))

ggplot(data = world) +
  geom_sf() +
  geom_text(data= world_points,aes(x=X, y=Y, label=name),
            color = "darkblue", fontface = "bold", size = 4,
            check_overlap = FALSE) +
  annotate(geom = "text", x = -90, y = 26, label = "Gulf of Mexico", 
           fontface = "italic", color = "grey22", size = 4) +
  coord_sf(xlim = c(-102.15, -74.12), ylim = c(7.65, 33.97), 
           expand = FALSE)


#7th, put everything together
ggplot(data = world) +
  geom_sf(fill= 'antiquewhite') +
  geom_text(data= world_points,aes(x=X, y=Y, label=name),
            color = 'darkblue', fontface = 'bold', 
            check_overlap = FALSE) +
  annotate(geom='text',x=-90,y=26,label='Gulf of Mexico',
           fontface = 'italic', color = 'grey22', size = 6) +
  annotation_scale(location = 'bl', width_hint = 0.5) +
  annotation_north_arrow(location = 'bl', which_north = 'true', 
          pad_x = unit(0.75, 'in'), pad_y = unit(0.5, 'in'),
          style = north_arrow_fancy_orienteering) +
  coord_sf(xlim = c(-102.15, -74.12), ylim = c(7.65, 33.97), 
           expand = FALSE) +
  xlab('Longitude') + ylab('Latitude') + 
  ggtitle('Map of the Gulf of Mexico and the Caribbean Sea') +
  theme(panel.grid.major = element_line(color = gray(.5),
        linetype = 'dashed', size = 0.5), 
        panel.background = element_rect(fill = 'aliceblue'))


#Finally, export the map 
ggsave("C:/Matias/Cursos/2019_Mexico/Plots_Maps/map.pdf")
ggsave("C:/Matias/Cursos/2019_Mexico/Plots_Maps/map_web.png",
       width = 6, height = 6, dpi = "screen")
#ggsave saves that last map on display....


#What if we want to add some field sites (e.g. point data)?

#create the site data
(sites <- data.frame(longitude = c(-80.144005, -80.109), 
                     latitude = c(26.479005, 26.83)))

#quickest way to add points to map
ggplot(data = world) +
  geom_sf() +
  geom_point(data = sites, aes(x = longitude, y = latitude),
             size = 4, shape = 23, fill = "darkred") +
  coord_sf(xlim = c(-88, -78), ylim = c(24.5, 33), 
           expand = FALSE)

#more flexible way... 
#  use 'sf' to convert sites to a coordinate system 
#   (e.g. more useful if changing projections)
(sites <- st_as_sf(sites, coords = c("longitude", "latitude"), 
                   crs = 4326, agr = "constant"))
ggplot(data = world) +
  geom_sf() +
  geom_sf(data = sites, size = 4, shape = 23, 
          fill = "darkred") +
  coord_sf(xlim = c(-88, -78), ylim = c(24.5, 33), 
           expand = FALSE)


#Add polygon data (e.g. States)
library("maps")   #only useful if plotting USA stuff...
              # used here for illustrative purposes...
states <- st_as_sf(map("state", plot = FALSE, fill = TRUE))
head(states)

#get the centroid for plotting each state name
states <- cbind(states, st_coordinates(st_centroid(states)))
ggplot(data = world) +
  geom_sf() +
  geom_sf(data = states, fill = NA) + 
  geom_text(data = states, aes(X, Y, label = ID), size = 5) +
  coord_sf(xlim = c(-88, -78), ylim = c(24.5, 33), 
           expand = FALSE)

#move state name a bit
states$nudge_y <- -1
states$nudge_y[states$ID == "Florida"] <- 0.5
states$nudge_y[states$ID == "South Carolina"] <- -1.5
ggplot(data = world) +
  geom_sf() +
  geom_sf(data = states, fill = NA) + 
  geom_label(data = states, aes(X, Y, label = ID),size = 5,
             fontface = "bold", nudge_y = states$nudge_y) +
  coord_sf(xlim = c(-88, -78), ylim = c(24.5, 33),
           expand = FALSE)


#highlight different counties (polygons...)
counties <- st_as_sf(map("county", plot = FALSE,
                         fill = TRUE))
counties <- subset(counties, grepl("florida", counties$ID))
counties$area <- as.numeric(st_area(counties))

#add counties as a gray outline
ggplot(data = world) +
  geom_sf() +
  geom_sf(data = counties, fill = NA, color = gray(.5)) +   
  coord_sf(xlim = c(-88, -78), ylim = c(24.5, 33),
           expand = FALSE)

#show counties by some property 
#  (e.g. area; in our case, some random variable)
Area=runif(length(counties$ID),1,10)
ggplot(data = world) +
  geom_sf() +
  geom_sf(data = counties, aes(fill = Area)) +
  scale_fill_viridis_c(trans = "sqrt", alpha = .4) +
  coord_sf(xlim = c(-88, -78), ylim = c(24.5, 33),
           expand = FALSE)


#Show more specific data points
flcities <- data.frame(
  state = rep("Florida", 5), 
  city = c("Miami","Tampa", "Orlando", "Jacksonville", "Sarasota"),
  lat = c(25.7616798,27.950575, 28.5383355, 30.3321838, 27.3364347),
  lng = c(-80.1917902,-82.4571776, -81.3792365, -81.655651, -82.5306527))

(flcities <- st_as_sf(flcities,coords=c("lng","lat"),remove=FALSE, 
                      crs = 4326, agr = "constant"))
ggplot(data = world) +
  geom_sf() +
  geom_sf(data = counties, fill = NA, color = gray(.5)) +
  geom_sf(data = flcities) +
  geom_text(data = flcities, aes(x = lng, y = lat,
        label = city), size = 3.9, col = "black", fontface = "bold") +
  coord_sf(xlim = c(-88, -78), ylim = c(24.5, 33), expand = FALSE)

#...doesn't look that good, 

# let's move the legend..
library("ggrepel")
 
ggplot(data = world) +
  geom_sf() +
  geom_sf(data = counties, fill = NA, color = gray(.5)) +
  geom_sf(data = flcities) +
  geom_text_repel(data = flcities, aes(x = lng, y = lat,
                  label = city),fontface = "bold", 
                  nudge_x = c(1, -1.5, 2, 2, -1),
                  nudge_y = c(0.25,-0.25, 0.5, 0.5, -0.5)) +
  coord_sf(xlim = c(-88, -78), ylim = c(24.5, 33), 
           expand = FALSE)

```

<br><br>

# Another nice approach: tmap

source:

https://geocompr.robinlovelace.net/adv-map.html

```{r}
library(tmap)
library(spData)

#Just like ggplot2, tmap is based on the idea of 
#  a ‘grammar of graphics’ 
#This involves a separation between the input data 
#  and the aesthetics (how data are visualised): each
#  input dataset can be ‘mapped’ in a range of 
#  different ways including location on the map (defined by 
# data’s geometry), color, and other visual variables.


# Add fill layer to nz shape
tm_shape(nz) +
  tm_fill() 

# Add border layer to nz shape
tm_shape(nz) +
  tm_borders() 

# Add fill and border layers to nz shape
nz=tm_shape(nz) +
  tm_fill() +
  tm_borders() 
nz

```
many more useful features, including how to animate maps..

...check here:

https://geocompr.robinlovelace.net/adv-map.html


<br><br>


# Mapview: interactive maps
```{r}
library(mapview)
library(leaflet)
m=mapview(breweries)

m2=leaflet() %>% 
  addProviderTiles("OpenStreetMap") %>%
  addExtent(breweries) 

#save as html
mapshot(m, url = paste0('C:/Matias/Cursos/2019_Mexico/Plots_Maps',
                        "/map.html"))
mapshot(m2, url = paste0('C:/Matias/Cursos/2019_Mexico/Plots_Maps',
                         "/map2.html"))


#can also use tmap
tmap_mode("view")
nz

```
<br>

# Wrap up exercise.
Try to recreate the map of Western Australia shown