---
title: "Part 2. data frames, summary statistics, plotting of frequency data"
output: 
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: yes
    theme: united
---

<style type="text/css">
h1.title {
  font-size: 30px;
    color: Maroon;
  text-align: center;
          }
h3.subtitle {
  font-size: 22px;
    color: Maroon;
  text-align: center;
            }
h4.author { 
    font-size: 24px;
      color: FireBrick;
  text-align: center;
          }
h4.date { 
  font-size: 18px;
  text-align: center;
        }
h1{
  font-size: 22px;
  color: DarkBlue;
  }
h2{
  font-size: 20px;
  color: Blue;
  }
h3{
  font-size: 18px;
  color: SteelBlue;
  }
body{
    font-family: Helvetica;
    font-size: 14pt;
    }
code.r{
  font-size: 16pt;
      }
pre {
  font-size: 16pt;
}
</style>


```{r globaloptions, include=FALSE}
knitr::opts_chunk$set(fig.width = 6,fig.height = 6,
                        echo = TRUE, warning=FALSE,message=FALSE)
```

# Lesson goals
We're going to create data frames and do some plotting.

```{r }
#make sure to start with a clean slate, with no junk in memory 
rm(list=ls())

# Create some vectors
ObsAge <- c(6,1,9,3,7,4,3,5,1,4)
ObsLen <- c(198,53,289,136,229,194,142,166,51,158)
length(ObsAge)==length(ObsLen)      #same length
N_Observations <- length(ObsAge)    #let's store this  
(N_Observations <- length(ObsAge))  #a way to print without printing

plot(x=ObsAge,y=ObsLen)
plot(ObsAge,ObsLen)   #same

#let's improve the plot (use additional arguments, type ?plot)
plot(ObsAge,ObsLen,frame.plot=F,cex=0.6,pch=16)

# what other symbol types are there? type ?pch

# redraw & change axis labels, increase point size, change color
plot(ObsAge, ObsLen, frame.plot=F, cex=2, pch=16, ylab="Total length, mm", xlab="Age class, y",col="steelblue",cex.axis=1.25,cex.lab=1.5)
```
<br> <br>
some R color options
http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf

```{r }
# redraw and change the range for the x and y axes
max(ObsLen)
max(ObsAge)
plot(ObsAge, ObsLen, frame.plot=F, cex=2, pch=16, ylab="Total length, mm", xlab="Age class, y",col="steelblue",cex.axis=1.25,cex.lab=1.5,
     xlim=c(0,20), ylim=c(0,400))

#be a bit smarter with axis limits...also change axis orientation
plot(ObsAge, ObsLen, frame.plot=F, cex=2, pch=16, ylab="Total length, mm", xlab="Age class, y",col="steelblue",cex.axis=1.25,cex.lab=1.5,las=1,
     xlim=c(min(ObsAge),max(ObsAge)), ylim=c(min(ObsLen),max(ObsLen)))

#Exercise: try las =0, =1, =2, =3
#          try changing points to lines (hint: search ?par 'type')
#           try changing colors (points, axis, labels etc)

```

## Introducing data frames
+ An excel spreadsheet often contains multiple columns and rows of data.
+ We can store our data this way in R (but better...)
+ Data frames, unlike matrices (we'll touch on those later), can combine numeric, character and factor (i.e. a character with repeated levels) variables 

```{r }
#Let's create a data frame containing the above age and length data
FishDat <- data.frame(ObsAge, ObsLen)
FishDat

head(FishDat)  #the data frame header
head(FishDat,n=2)
FishDat[1:2,]  #same 

tail(FishDat,n=2)   #the last 2 records

FishDat[,1]  #access columns
FishDat[,"ObsAge"]  #same
FishDat$ObsAge      #same


FishDat[1:10,1]  # first 10 ages in the data set
FishDat$ObsAge[1:10] #same

FishDat <- data.frame(Age=ObsAge, Length=ObsLen) #different names
head(FishDat,3)
colnames(FishDat)=c("ObsAge","ObsLen")  #rename it

N_lengths <- length(FishDat$ObsLen) # how many lengths do we have?
N_lengths
nrow(FishDat)
dim(FishDat)  #data frame dimensions

FishDat$FishNum <- seq(from=1, to=N_lengths, by=1) #add a new column
dim(FishDat)

head(FishDat,3)
FishDat <- FishDat[,c(3,1,2)] #re arrange columns
head(FishDat,3)
FishDat <- FishDat[,c("FishNum","ObsAge","ObsLen")] #same
head(FishDat,3)

RowForFish3 <- which(FishDat$FishNum==3)  #access a specific record
RowForFish3
FishDat$ObsLen[which(FishDat$FishNum==3)]  # get its length
FishDat[RowForFish3,3]                     #same
FishDat$ObsLen[RowForFish3]                #same 

#Let's assume there's an error in the database and we need to fix it
#(better to fix errors in code rather than in original data set)
original=FishDat$ObsLen[RowForFish3]
FishDat[RowForFish3,]
FishDat$ObsLen[RowForFish3]=999
FishDat[RowForFish3,]
FishDat$ObsLen[RowForFish3]=original
FishDat[RowForFish3,]
#plot data using the data frame
plot(FishDat$ObsAge, FishDat$ObsLen, frame.plot=F, cex=2, pch=16, ylab="Total length, mm", xlab="Age class, y",col="steelblue",cex.axis=1.25,cex.lab=1.5,las=1,
     xlim=c(min(ObsAge),max(ObsAge)), ylim=c(min(ObsLen),max(ObsLen)))

```

## Some more coding & how to calculate & plot frequency distributions
```{r }
dummy=round(rnorm(1000,50,10)) #first create some dummy length data                       (don't worry about these functions for the moment)
LenFreq <- table(dummy) #create a table of age observations
plot(LenFreq)   # Let's plot it
plot(LenFreq,type="o", pch=16, lwd=0.8,las=1,
     xlab="Length class",ylab="Frequency")   # make it a bit better
barplot(LenFreq,xlab="Length class",ylab="Frequency",las=1) # same

# let's change the colour
barplot(LenFreq,xlab="Length class",ylab="Frequency", col="light blue",las=1)
barplot(LenFreq,xlab="Length class",ylab="Frequency", col=rainbow(length(LenFreq)), las=1)

#create different length categories
sort(unique(dummy))
max(dummy)
Max <- 10*ceiling(max(dummy)/10)
Min <- 10*floor(min(dummy)/10)
LenInterval <- 10  #length interval (every 10 cm)
LenCats <- seq(from=Min, to=Max, by=LenInterval) #length categories

FreqDist <-  table(cut(dummy, breaks=LenCats, right=FALSE))
FreqDist   #type ?cut
#compare to
table(cut(x=dummy, breaks=LenCats))
#it has changed, because the right cut-offs have changed

# To plot, we can use the hist() function
hist(dummy, breaks=LenCats, right=FALSE, col="light blue", main="", xlab="10 mm Length category", ylab="Frequency",las=1)
```
```{r }
# We can use hist() to create the table (set it equal to a new object)
HistData=hist(dummy, breaks=LenCats, right=FALSE, col="light blue", main="", xlab="10 mm Length category", ylab="Frequency",las=1)
HistData$breaks  #can acces the histogram
HistData$counts

#if don't need to plot the histogram
HistData=hist(dummy, breaks=LenCats, right=FALSE, plot=FALSE)

```

# Wrap up exercise
1. Create 3 vectors with numbers of the same length, using numbers ranging from 0 to 50
2. Join the vectors together to create a data frame
3. Make a scatter plot, with one variable (column of the data frame) plotted against the other
4. Take one of the variables (one column), work out the minimum and maximum size.
5. Determine a length interval, define some length classes, calculate the frequency  with which numbers for the variable of interest fall into your defined length categories and plot.



```{r }

ObsAge <- c(4,5,3,12,8,17,8,9,8,6,11,12,8,10,11)
ObsAge

ObsLen <- c(524,590,491,702,645,707,680,640,676,665,602,620,672,692,666)
ObsLen

length(ObsAge)
length(ObsLen)

FishDat <- data.frame(ObsAge,ObsLen)
FishDat

max(ObsAge)
max(ObsLen)

plot(FishDat$ObsAge, FishDat$ObsLen,frame.plot=F, cex=0.6, cex.lab=1.2, pch=16,las=1, 
     ylab="Total length, mm", xlab="Age class, y", xlim=c(0,20),ylim=c(0,750))

table(FishDat$ObsAge)

AgeFreq <- table((FishDat$ObsAge))

AgeClasses <- seq(1,17,1)

MaxLen <- 750

LenInterval <- 20

LenCats <- seq(from=460, to=MaxLen, by=LenInterval)
LenCats

FreqDist <-  table(cut(FishDat$ObsLen, breaks=LenCats, right=FALSE))
FreqDist

hist(FishDat$ObsLen, breaks=LenCats, right=FALSE, col="light blue", main="", xlab="20 mm Length category", ylab="Frequency",las=1)

```