---
title: "Aplicación del paquete estadístico R para la visualización y el análisis de datos              pesqueros y ecológicos"
subtitle: "Unidad Académica Mazatlán"
author: " <br> Matias Braccini <br> Department of Primary Industries and Regional Development  <br> Western Australia"
date: "7-11 October 2019"
output: 
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: yes
    theme: united
---

<style type="text/css">
h1.title {
  font-size: 30px;
    color: Maroon;
  text-align: center;
          }
h3.subtitle {
  font-size: 22px;
    color: Maroon;
  text-align: center;
            }
h4.author { 
    font-size: 24px;
      color: FireBrick;
  text-align: center;
          }
h4.date { 
  font-size: 18px;
  text-align: center;
        }
h1{
  font-size: 22px;
  color: DarkBlue;
  }
h2{
  font-size: 20px;
  color: Blue;
  }
h3{
  font-size: 18px;
  color: SteelBlue;
  }
body{
    font-family: Helvetica;
    font-size: 14pt;
    }
code.r{
  font-size: 16pt;
      }
pre {
  font-size: 16pt;
}
</style>


```{r globaloptions, include=FALSE}
knitr::opts_chunk$set(fig.width = 12,fig.height = 8,
                        echo = TRUE, warning=FALSE,message=FALSE)
```
<br>
GitHub repository: https://github.com/JuanMatiasBraccini/Git_R_course <br>
Dropbox repository: https://www.dropbox.com/home/R_course_Mazatlan_2019

#Acknowledgements
Big thanks to Simon De Lestang, Ainslie Denham and Alex Hesp for providing with reference code and material 

# Introduction to R and Rstudio
1.  A script is a combination of **comments** and **words** which is R syntax (code)
```{r }
#calculate mean
mean(1:10)
```
2.  "R studio" provides an interface for using the program R and it has 4 main panes:
+ top left is where you write code (this is where you save your script)
+ below is the console where you see outputs and code that you have run
+ top right window shows info on objects 
+ bottom right window shows plot, help, etc.

##Using R like a calculator
```{r }
# what is 4 + 5?
#type 4+5 and press 'Run' or 'ctrl+Enter' (this is how we run code)
4 + 5

# Multiplication, let's try 21 * 6
21 * 6

# Subtraction
63 - 2

# Division
10 / 5

# A mixture of basic operations
2 * (10 + 6)
```
## Various ways to run code
+ To run a single line of code, position the cursor and press 'Run' or 'Ctrl + Enter'
+ To run several lines of code at once, highlight them all and press 'Run' or 'Ctrl + Enter'
+ To run whole code at once, 'Ctrl + a' and 'Run'


## Storing values as "variables"
```{r}

 # Store information in computer memory, so that we can use it later
FirstNumber <- 4    
SecondNumber <- 5
#note: when naming variables use names that describe what information is being stored, so that the name means something to you
Result=FirstNumber+SecondNumber #can use '<-' or '='
Result   #the new variable (where the outcome is stored)
# In the top right window (environment window), you should see under 'Values' our variable names
# So the computer now knows that our first variable, FirstNumber, is equal to 4, and our second variable, SecondNumber, is equal to 5

#note that R is case-sensitive, type 'firstnumber'

#if not interested in keeping a variable in the memory
rm(FirstNumber)
# so it's now gone from the environment window.

#if we need to clear the entire memory
rm(list=ls()) #if we put this at the beginning of a script, we start with no info being stored in memory

# Text variables (called 'strings')
CalendarMonth <- "Jan"   #note how we need to add " " around the text
CalendarMonth
```

## Storing several values as a vector
```{r}
# Often we want to store, manipulate, etc., multiple numbers at once
# For example, we can store several numbers in a "vector".
NumberVector1 <- c(1,2,3,4,5)
NumberVector1    #look at the top right pane...

#vector calculations
NumberVector2 <- c(6,7,8,9,10)
NumberVector1 + NumberVector2  #(element wise)
NumberVector1 * NumberVector2

#joining vectors
NumberVector3 <- c(NumberVector1,NumberVector2)

# what about text
TextVector1 <- c("Jan","Feb","Mar")
TextVector2 <- c("Apr","May","Jun")
TextVector3=c(TextVector1,TextVector2)  #type TextVector1+TextVector2
```

## Specifying number sequences
```{r}
# how can we create the numbers 1,2,.., 10, without writing all these numbers?
seq(from=1,to=10,by=1)
# this is the same as
seq(1,10,1)  # seq is an inbuilt function (it has arguments); If you don't know the function/arguments: ask Google, or type ?seq

#Let's store these numbers
NumberVector4 <- seq(1,10,1)
NumberVector4

#Exercise: create a sequence of numbers starting at 0, finishing            at 100, and incrementing by 2

#Let's reverse the sequence
seq(10,1,-1)
rev(seq(1,10,1)) #note: there are multiple ways of getting to the          same solution....

#Exercise: create a sequence of numbers increasing from 0 to 20 by 2            and then decreasing back to 0 by 2 without repeating 20 

```

## Basic summary statistics
```{r}
NumberSequence <- c(seq(0,20,2),seq(18,0,-2))
length(NumberSequence)        #how namy numbers in our sequence?
max(NumberSequence)           #what is the maximum value?
min(NumberSequence)           #what is the minimun value?
range(NumberSequence)         #what is the range?
sum(NumberSequence)           #what do all the numbers add up to?
mean(NumberSequence)          #what is the mean?
sd(NumberSequence)            #what is the standard deviation?
median(NumberSequence)        #what is the 50th percentile (median) 
#...type percentile(NumberSequence)
quantile(NumberSequence,0.75) #what is the 75th percentile
summary(NumberSequence)       #summary

```
## Basic functions cheat sheet
https://www.rstudio.com/wp-content/uploads/2016/10/r-cheat-sheet-3.pdf
>br>

## Accessing elements within a vector (subsetting)
```{r}
NumberSequence[1] # the first number of that sequence
NumberSequence[5] # the fifth number of that sequence
NumberSequence[1:5] # the first five numbers of that sequence
NumberSequence[c(1,5)] # the first and fifth numbers of that sequence

#what if we want to find a particular number in the vector?
id=which(NumberSequence==4) #the == indicates a logical operation
id
NumberSequence[id]

#find the values NOT EQUAL to 4
id=which(!NumberSequence==4) 
NumberSequence[id]

#which ones had a value less than or equal to 10?
id=which(NumberSequence<=10)
NumberSequence[id]

#which ones had a value more than 10?
id=which(NumberSequence>10)
NumberSequence[id]

# which ones are less/equal to 4 OR  greater/equal to 18?
id=which(NumberSequence<=4 | NumberSequence>=18)
NumberSequence[id]

#which ones are less/equal to 10 AND greater/equal to 4?
id=which(NumberSequence<=10 & NumberSequence>=4)
NumberSequence[id]

#logical operations for character vectors
which(TextVector1=="Feb")
which(TextVector1%in%c("Feb","Mar"))
id=which(TextVector1%in%c("Feb","Mar"))
TextVector1[id]
TextVector1[which(TextVector1%in%c("Feb","Mar"))]
which(month.abb=="Jun")     # vector already defined in R
which(month.name=="April")  # vector already defined in R

```

## vectors of repeated numbers
```{r}
rep(1,5)
rep(2,3)
c(rep(1,5),rep(2,3))  #combined results
rep(1:2, each=2)     #repeat each times

```

## Wrap up exercise
1. Create a vector containing the numbers 1 to 10 
2. Create a second vector containing the numbers 1 to 5
3. Join them together
4. Add 5 to each number, then multply the resultant numbers by 6
5. Calculate the minimum, maximum and mean of the resultant set of numbers
6. Work out which of the numbers are greater then 60 
7. Subtract 10 from only those numbers that are greater than 60, and add them all up!

```{r}
NumberVector1 <- 1:10
NumberVector1
NumberVector2 <- 1:5
NumberVector2
NumberVector3 <- c(NumberVector1,NumberVector2)
NumberVector3
(NumberVector3+5)*6 -> NumberVector4  #either way works...
NumberVector4
summary(NumberVector4)
id=which(NumberVector4>60)
id
NumberVector4[id]
sum(NumberVector4[id]-10)
```