---
title: "Lesson 8. Demography and population dynamics"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "C:/Matias/Cursos/2019_Mexico/htmls") })

output: 
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: yes
    theme: united
---

<style type="text/css">
h1.title {
  font-size: 30px;
    color: Maroon;
  text-align: center;
          }
h3.subtitle {
  font-size: 22px;
    color: Maroon;
  text-align: center;
            }
h4.author { 
    font-size: 24px;
      color: FireBrick;
  text-align: center;
          }
h4.date { 
  font-size: 18px;
  text-align: center;
        }
h1{
  font-size: 22px;
  color: DarkBlue;
  }
h2{
  font-size: 20px;
  color: Blue;
  }
h3{
  font-size: 18px;
  color: SteelBlue;
  }
body{
    font-family: Helvetica;
    font-size: 14pt;
    }
code.r{
  font-size: 16pt;
      }
pre {
  font-size: 16pt;
}
</style>


```{r globaloptions, include=FALSE}
knitr::opts_chunk$set(fig.width = 6,fig.height = 6,error=TRUE,
                        echo = TRUE, warning=FALSE,message=FALSE)
```

# Lesson goals
We're going to develop and fit Leslie matrices and surplus production population dynamics models.

# Leslie matrices (Wikipedia)

- The Leslie matrix is one of the most well known ways to describe the growth of populations (and their projected age distribution), in which a population is closed to migration, growing in an unlimited environment, and where only one sex, usually the female, is considered
- It is a square matrix (i.e. same number of rows and columns)
- At each time step, the population is represented by a vector $N_t$ with an element for each class (age or size) where each element indicates the number of individuals currently in that class
- The key elements are:
  - $n_x$, the count of individuals (n) of each class x
  - $s_x$, the fraction of individuals that survives from class x to class x+1,
  - $f_x$, fertility at class x, the number of offspring produced at the next class weighted by the probability of reaching the next class.

$$N_{t+1}=X N_t$$
$$\mathbf{X} = \left[\begin{array}
{rrr}
f_1 & f_2 & f_3 & f_4 & f_5\\
s_1 & 0 & 0 & 0 & 0\\
0 & s_2 & 0 & 0 & 0\\
0 & 0 & s_3 & 0 & 0\\
0 & 0 & 0 & s_4 & 0\\
\end{array}\right]
$$
$$\mathbf{N_t} = \left[\begin{array}
{rrr}
n_1 \\
n_2 \\
n_3 \\
n_4 \\
n_5 \\
\end{array}\right]
$$

## Let's fit a Leslie matrix for white sharks

source: Braccini et al 2017
(https://www.sciencedirect.com/science/article/pii/S0304380017303447)

```{r}
library(triangle)
library(popbio)     #for solving matrices


#1. Biological data
first.age=0   #first age class, assuming post-breeding census 
              # (this is the 0-1 age class actually)
RangeA=c(40,60)  #max age range
k.mean=0.056       #growth parameters
Linf.mean=7.19
Lo.mean=1.4
Range.reprod_cycle=1:3   # possible reproductive cycles
prob.rep.cycle=c(0.05,0.25,0.7)   # probs for 1, 2 or 3-year reprod cycle
Rangefec=2:10         #litter size
age.mat=c(9,17)      #age at maturity of females
Aver.T=14         #average water temperature in WA
sd.Aver.T=1.5
bwt=7.5763e-06;awt=3.0848 #length-weight parameters
Dry.w=0.2   #convertion wet to dry weight (Cortes 2002)
sexratio=0.5  #pups sex ratio


#2.  Some useful functions
# Natural mortality
M.fun=function(A,k,Linf,Aver.T,age.mat,bwt,awt,W.G,W.noG)
{
  #STEP 1. calculate M from different methods (see Kenchington 2013)
  #.age invariant
  #Jensen (1996)
  m.Jensen.1=1.5*k
  m.Jensen.1=rep(m.Jensen.1,length(age))
  
  m.Jensen.2=1.65/age.mat
  m.Jensen.2=rep(m.Jensen.2,length(age))
  
  #Pauly (1980)  
  m.Pauly=10^(-0.0066-0.279*log10(Linf*100)+0.6543*log10(k)+0.4634*log10(Aver.T))
  m.Pauly=rep(m.Pauly,length(age))
  
  #Hoenig (1983), combined teleost and cetaceans    
  m.Hoenig=exp(1.44-0.982*log(A))      
  m.Hoenig=rep(m.Hoenig,length(age))
  
  #Then et al (2015)
  m.Then.1=4.899*A^(-0.916)
  m.Then.1=rep(m.Then.1,length(age))
  #m.Then.2=4.118*(k^0.73)*(Linf^(-0.33))
  
  #.age dependent
  #Peterson and Wroblewski 1984 (dry weight in grams)
  wet.weight=(7.5763e-06)*fork^3.0848 
  m.PetWro=1.92*((wet.weight*Dry.w)*1000)^-0.25
  
  #Lorenzen 1996 (weight in grams)
  m.Lorenzen=3*(wet.weight*1000)^-0.288
  
  #Gislason et al (2010) (lengths in grams)
  m.Gislason=1.73*(mid.FL.fem^-1.61)*((Linf*100)^1.44)*k
  if(m.Gislason[1]>1)m.Gislason=rep(NA,length(age))
  
  
  #STEP 2. get mean at age
  nat.mort=data.frame(m.Jensen.2,m.Pauly,m.Hoenig,m.Then.1,m.PetWro,m.Lorenzen,m.Gislason)
  N=ncol(nat.mort)
  nn=length(age)
  nat.mort.all=nat.mort
  
  Weights=data.frame(W.J=rep(W.noG,nn),W.P=rep(W.G,nn),W.H=rep(W.noG,nn),W.T=rep(W.noG,nn),
                     W.P=rep(W.noG*1,nn),W.L=rep(W.noG*1,nn),W.G=rep(W.G*1,nn))
  
  nat.mort=data.frame(nat.mort,Weights)
  nat.mort=apply(nat.mort, 1, function(x) weighted.mean(x[1:N], x[(N+1):ncol(nat.mort)],na.rm = T))
  
  
  return(nat.mort)
}

# Fill ages for square matrix
add.missing.age=function(data,A)
{
  test=list()
  for(i in 1:length(data))
  {
    datos=data[[i]]
    if(length(datos)<A)
    {
      extra=A-length(datos)
      datos=c(datos,rep(NA,extra))
    }
    test[[i]]=datos
    if(length(datos)>A)test[[i]]=datos[1:A]
  }
  dummy=do.call(rbind,test)
  return(dummy)
}


#3. Leslie matrix
set.seed(999)
#Draw random samples of parameters and calculate population quantities
Ns <- 1000	  #number of simulations
iterations=1:Ns
mx.sim=m.SIM=Proyec.matrix=vector("list",length = Ns)
A.SIM=vector(length = Ns)
for (s in iterations)
{
  
  #1. Draw parameter samples and build functions
    #maximum age
  A.sim=floor(rtriangle(1,
                        a=RangeA[1],
                        b=RangeA[2]+1,
                        c=RangeA[1])) #triangular dist
  A.SIM[s]=A.sim 
        
    #number of age classes
  age=first.age:A.sim
        
    #reproductive cycle
  Reprod_cycle.sim=sample(Range.reprod_cycle,1,replace=T,
                          prob = prob.rep.cycle) 
  Pmax.sim=1/Reprod_cycle.sim
        
    #fecundity and age at maturity
  Meanfec.sim=rep(ceiling(rtriangle(1,
                  a=Rangefec[1]-1,
                  b=Rangefec[length(Rangefec)],
                  c=Rangefec[length(Rangefec)])),length(age)) #triangular dist
  age.mat.sim=round(rtriangle(1,
                  a=age.mat[1],
                  b=age.mat[2],
                  c=floor(mean(age.mat))))    #triangular dist
   
    #temperature
  Aver.T.sim=rnorm(1,Aver.T,sd.Aver.T)    #normal dist
        
    #total length
  total.length=Lo.mean+(Linf.mean-Lo.mean)*(1-exp(-k.mean*age))
        
    #fork length
  fork=(0.9442*total.length*100)-5.7441 
  mid.FL.fem=fork
        
    #natural mortality
  m.sim=M.fun(A=A.sim,k=k.mean,Linf=Linf.mean,Aver.T=Aver.T.sim,
              age.mat=age.mat.sim,bwt,awt,W.G=1,W.noG=1)
  m.SIM[[s]]=m.sim
  
    #survivorship
  S=exp(-m.sim)         
        
    #proportion surviving
  lx=rep(NA,length(age))
  lx[1]=1.0
  for (i in 2:(length(age)))lx[i]=lx[i-1]*S[i]
        
    #reproductive schedules   
  MF=c(rep(0,(age.mat.sim-1)),Meanfec.sim[age.mat.sim:length(Meanfec.sim)])
  mx=MF*sexratio*Pmax.sim
  mx.sim[[s]]=mx
  
    #probability of surviving (for birth-pulse, post-breeding census)
  px=vector(length=length(lx))
  for(i in 2:length(lx)) px[i-1]=(lx[i])/(lx[i-1])
  
    #fertility  (for birth-pulse, post-breeding census)
  bx=mx*px

    #projection matrix   
  PX=px
  PX=PX[-length(PX)]
  BX=bx
  n=length(BX)
  Data=matrix(0,nrow=n,ncol=n)
  diag(Data)[-nrow(Data)]=PX
  Data=rbind(matrix(0,nrow=1,ncol=n),Data)
  Data=Data[-(n+1),]
  Data[1,]=BX
  rownames(Data)=colnames(Data)=(first.age+1):n
  Proyec.matrix[[s]]=Data
}

#Solve matrix
LAMBDA=sapply(Proyec.matrix,lambda)
r.sim=log(LAMBDA)         
    
#Calculate quantities of interest
t2.sim=log(2)/r.sim           #pop doubling time
v=sapply(Proyec.matrix,reproductive.value)    #reproductive value
Elasticities=sapply(Proyec.matrix,elasticity) #elasticities
 
   
#4. Extract average elasticities per life stage
names.elast.stage=c("juv.surv","adult1.surv","adult2.surv",
                    "juv.fec","adult1.fec","adult2.fec") 
n.elas.stages=length(names.elast.stage)
Elast.stage=matrix(NA,ncol=n.elas.stages,nrow=length(Elasticities))
nn=1:n.elas.stages
juv=(first.age):(age.mat[1]-1)
adul1=age.mat[1]:age.mat[2]
adul2=(age.mat[2]+1):RangeA[2]
for (e in 1:length(Elasticities))
{
      test=Elasticities[[e]]
      Elast.stage[e,nn[1]]=sum(test[2:nrow(test),juv])
      Elast.stage[e,nn[2]]=sum(test[2:nrow(test),adul1])
      Elast.stage[e,nn[3]]=sum(test[2:nrow(test),adul2[adul2<=A.SIM[e]]])
      
      Elast.stage[e,nn[4]]=sum(test[1,juv])
      Elast.stage[e,nn[5]]=sum(test[1,adul1])
      Elast.stage[e,nn[6]]=sum(test[1,adul2[adul2<=A.SIM[e]]])
    }
colnames(Elast.stage)=names.elast.stage 
    
    
#5 Fill in missing age classes 
mx.sim=add.missing.age(mx.sim,RangeA[2])
m.SIM=add.missing.age(m.SIM,RangeA[2])
v.SIM=add.missing.age(v,RangeA[2])

      
#6. Remove  any param combos that yields negative growth
LAMBDA.reject=LAMBDA[LAMBDA<1]  #store negative growth separetely
ID.lam.rej=which(LAMBDA<1)
if(length(ID.lam.rej)>0)
{
      LAMBDA=LAMBDA[-ID.lam.rej]        
      r.sim=r.sim[-ID.lam.rej]
      t2.sim=t2.sim[-ID.lam.rej]
      A.SIM=A.SIM[-ID.lam.rej]
      Elasticities=Elasticities[-ID.lam.rej]
      Elast.stage=Elast.stage[-ID.lam.rej,]
      Proyec.matrix=Proyec.matrix[-ID.lam.rej]
}
    
#7. Show basic results
  #Lambda, r, doubling time
par(mfcol=c(3,1),mar=c(4,2,1,1),oma=c(2,2,.1,.1),cex.lab=1.5,
    mgp=c(2.2,.6,0))
hist(LAMBDA,main='',col="orange",ylab="",
     xlab=expression(paste(lambda," " (year^-1),sep="")))
hist(r.sim,main='',col="orange",ylab="",
     xlab=expression('r '(year^-1)))
hist(t2.sim,main='',col="orange",ylab="",xlab="years")
mtext("Frequency",2,outer=T)

  #Mortality at age
m.med=apply(m.SIM,2,median,na.rm=T)
m.upper=apply(m.SIM,2,quantile,probs=.975,na.rm=TRUE)
m.lower=apply(m.SIM,2,quantile,probs=.025,na.rm=TRUE)
par(mfcol=c(1,1),mar=c(4,4,.1,.1),cex.lab=1,oma=c(4,4,.1,.1))

plot(1:RangeA[2],m.med,pch=19,col=1,cex=1.25,xlab="Age",
     ylab=expression("Natural mortality  " (year^-1)),
     ylim=c(0,max(m.upper)))
segments(1:RangeA[2],m.upper,1:RangeA[2],m.lower)


```

<br><br>

# Surplus production

Residual errors are assumed to be log-normally distributed so the corresponding negative log-likelihood
of the data is given by
$$l(I_t|\theta)=ln(\sigma)+\frac{1}{2}\ln(2\pi)+\frac{(ln(I_t)-ln(\hat{I}_t))^2}{\sigma^2}$$
where $I_t$ is the observed CPUE, $\hat{I}_t$ is the CPUE predicted by the model and $\sigma^2$ is the unknown variance of the observation or process error.

Given initial estimates for the unknown parameters $\theta=(r,k,\sigma^2,q)$, one can estimate
this unknown parameters by minimizing the negative log-likelihood

```{r}
#DATA SECTION
hake=read.table("C:/Matias/Cursos/2019_Mexico/Data.sets/Lesson8_Hake data.txt",header=T)
#Data obtained from Hilborn and Mangel (1997). Namibian Hake...

ct = hake$Catch/1000
cpue=hake$CPUE
year=hake$Year
iyr=cbind(year,year+0.2,year+0.4,year+0.55,year+0.65)
L = 3

#PARAMETER SECTION
theta=c(log.k=log(2.9),log.r=log(0.38),log.q=log(0.35),tau=7)

#PROCEDURE SECTION
M1=function(theta)
{
  with(as.list(theta),{
    bt=vector()
    k=exp(log.k); r=exp(log.r); q=exp(log.q)	#k, r and q in log space so thet cannot go negative when estimated by Optim
    bt[1]=k; n=length(ct)	
    for(i in 1:n)
    {
      bt[i+1]=bt[i]+r*bt[i]*(1-bt[i]/k)-ct[i]		#recursive loop to estimate biomass
    }
    epsilon=log(cpue)-log(q*bt[1:n])
    nloglike= -sum(dnorm(epsilon,0,1/tau,log=T))	#robust Optim. SD=1/tau, tau is precission, this avoids having a negative SD
    return(list(bt=bt,epsilon=epsilon,nloglike=nloglike,
                pred.cpue=q*bt[1:n],par=c(k=k,r=r,q=q,sig=1/tau)))		#par is the vector of parameters in the untransformed state
  })
}

#function prototypes
fn1 = function(theta)		M1(theta)$nloglike


#optimizations
f1 = optim(theta,fn1,method="BFGS",hessian=T)


#Compare parameter estimates
print(round(M1(f1$par)$par,2))


#Plot the residuals between the predicted and observed CPUE’s for each model
epsilon=cbind(M1(f1$par)$epsilon)
matplot(iyr[,1:1],epsilon,type="h",xlab="Year",ylab="Residuals")
legend("topright",paste("std for M",1:1," = ",round(sd(epsilon),2),sep=""),bty="n",lty=1,col=1)


#Show cpue fit and biomass
Stuff=M1(f1$par)
par(mfcol=c(2,1),mar=c(3,3,1,1),oma=c(1,2,.1,1),mgp=c(1.7,.6,0))
plot(cpue,ylab="CPUE")
lines(Stuff$pred.cpue)
plot(Stuff$bt,ylab="Biomass")


```

